Error Reporting Parsers: a Monad Transformer Approach

### 2013-08-05 article.tex
Edits and code fixups.
Signed-off-by: Edward Z. Yang <ezyang@mit.edu>

The goals of parsing are to: 1) decide whether a string is part of the language
in question, and 2) build a meaningful representation of the structure of the
parsed string.  In addition, practical parsers must recognize faulty input and
accurately report the location and cause of the problem.  Thus, a parser must
also decide what generates an error, what information is in the error, and how
error-reporting parsers compose.

@@

To get a feel for what parser combinators are and how they're used to
build parsers, we'll quickly create a minimal parser combinator library to see
the basic principles of their design before adding monad transformers to
the mix.  We'll divide the library into three main pieces of the code:

\begin{itemize}
 \item Parser datatype
 \item a primitive parser
 \item type class instances
 \item combinators
\end{itemize}

@@

We also want parsers to return a value, if they succeed.  So let's extend the return type
again to allow for a value, which also requires a second type parameter:
\begin{verbatim}
newtype Parser t a = 
    Parser {getParser :: [t] -> ([t], a)}
\end{verbatim}

However, this ignores the possibility of failure.  To allow for failure, the return value needs to
be wrapped in a Maybe:
\begin{verbatim}
newtype Parser t = 
    Parser {getParser :: [t] -> Maybe ([t], a)}
\end{verbatim}

@@

The \verb+<*>+ operator of Applicative runs parsers sequentially, 
and applies the result of the first to the result of the second; it
has to make sure that the (possibly modified) output token stream from
the first parser is passed to the second:
\begin{verbatim}  
import Control.Applicative  (Applicative(..))

@@

We can run parsers in sequence using the Applicative combinators.  First one parser
is run, then the second, and the token position is threaded between the two.
Note that both parsers must succeed in order for the combined parser to succeed:
\begin{verbatim}
*Main> getParser (item *> item) "abcde"
Just ("cde",'b')

@@

parseNest :: Parser Char [Nesting]
parseNest = many element
\end{verbatim}
Examples:
\begin{verbatim}
*Main> getParser parseNest "(((()))))"
Just (")",[Many [Many [Many [Many []]]]])

@@

*Main> getParser parseNest "(()abc(def)()"
Just ("(()abc(def)()",[])
\end{verbatim}
That concludes the implementation of a simple parser.
In the rest of this article, we'll rework and augment these basic parser 
combinators by generalizing the implementations and extending the result types,
while taking advantage of monad transformers to keep them modular.

@@

This is identical to the Parser type given before, except for the order 
of the result tuple, and so we can redefine the parser type as:
\begin{verbatim}
import Control.Monad.State       (StateT(..))

@@

There are two major benefits to our parsers from using monad transformers:
\begin{itemize}
 \item semantic building blocks -- the mtl library provides implementations
    for Maybe, Error, and State building blocks; using these, we can describe
    describe the semantics of our parsers, and
 \item type class instances -- in addition to providing the building blocks,
    the mtl provides instances for common type classes, so we only have to 
    worry about parser-specific operations.
\end{itemize}

@@

The simple language we'll use as a motivation for building parsers is Woof, a 
simple dialect of Lisp.  We'll be progressively adding features to a simple 
initial implementation on our way to creating an error-reporting parser.

The language definition in pseudo-BNF \cite{bnf} is:

@@

With the additional rule that whitespace and comments may appear in any 
amount before or after tokens.  Tokens are:
\begin{itemize}
  \item \verb+{+
  \item \verb+}+
  \item \verb+(+
  \item \verb+)+
  \item \verb+Symbol+
\end{itemize}

@@

\subsection{Preliminaries}
Our first Woof parser will be responsible for determining whether input
text conforms to the language definition and building an 
Abstract Syntax Tree (AST) representing the structure of the parsed input.
Here's the AST definition we'll use:
\begin{verbatim}
data AST
    = ASymbol String
    | ALambda [String] [AST]
    | ADefine String AST
    | AApp    AST  [AST]
  deriving (Show, Eq)
\end{verbatim}

We'll also want a function for running our parsers, using the type given in the
prevous section:
\begin{verbatim}
runParser :: Parser [t] a -> [t] -> Maybe (a, [t])
runParser = runStateT
\end{verbatim}

@@

Because we used MonadState's combinators to build \verb+item+, \verb+literal+ 
has a slightly different type; we also take advantage of \verb+satisfy+ to 
simplify its implementation:
\begin{verbatim}
literal :: (Eq t, MonadState [t] m, Alternative m) => t -> m t
literal c = satisfy ((==) c)
\end{verbatim}

@@

But what about the spec that said that comments and whitespace can occur 
anywhere before or after tokens -- how should that be handled?

First we need to define the whitespace and comment patterns -- and also rename
the \verb+some+ and \verb+many+ functions from Control.Alternative to more
clearly indicate their semantics:
\begin{verbatim}
many0 = many
many1 = some

@@

We want to throw away comments and whitespace, and let's throw them away after 
every token.  Let's do that using a combinator which first parses a token, and 
then throws away junk:
\begin{verbatim}
junk = many0 (whitespace <|> comment)

@@

With \verb+tok+ in hand, we fix \verb+opencurly+ and try it out:
\begin{verbatim}
opencurly = tok $ literal '{'

@@

Excellent!  The \verb+opencurly+ parser matches an open curly brace, failing
otherwise, and discard trailing whitespace and comments.

@@

The parser for our last token, Symbol, is more complicated because any number of
alphabetical characters is valid.  We'll use \verb+satisfy+ and a
predicate to check that a character is valid, and \verb+many1+ to 
match one or more valid characters:
\begin{verbatim}
symbol = tok $ many1 char
  where char = satisfy (flip elem (['a' .. 'z'] ++ ['A' .. 'Z']))

@@

The first combining form is function application.  The rule says that it's delimited
by matching parentheses, in between which must appear one form as the operator, followed
by any number of forms as the arguments to which the operator is applied.  We can
say that like so:
\begin{verbatim}
application =
    openparen    *>
    pure AApp   <*>
    form        <*>
    many0 form  <*
    closeparen
\end{verbatim}

@@

\section{Example 2: error-reporting}
While our first parser worked great on valid input, it wasn't helpful
when the input was malformed.  When the parser finds bad input, we'd like it to 
produce an informative error, indicating what and where the problem was.  
This allows a user to quickly find and correct problems.

@@

While our previous parsers had two different results -- Nothing and Just --
our new parsers have three possible results:  success, failure, and error. 
A parse failure means that a parser wasn't able to match the input, but
nothing bad happened; failures can be recovered from if there is an alternative
parse that succeeds.  However, an error means that an unrecoverable problem
has been found and that parsing should immediately stop; the parser will
not be able to recover, even if there are alternatives since backtracking
will not be performed.
Note that successful and failing parses are wrapped in an additional \verb+Right+
constructor:
\begin{verbatim}
*Main> runParser item "abc"
Right (Just ('a',"bc"))

>>>>

The goals of parsing are to: 1) decide whether a string is part of the language
in question, and 2) build a meaningful representation of the structure of the
parsed string.  In addition, practical parsers must recognize faulty input and
accurately report the location and cause of the problem.  Thus, a parser must
also decide what generates an error, what information is in the error and how
error-reporting parsers compose.

@@

To get a feel for what parser combinators are and how they're used to
build parsers, we'll first create a minimal parser combinator library.
Next, we'll add monad transformers to the mix.  We'll divide the library
into four main pieces:

\begin{itemize}
 \item the Parser datatype,
 \item a primitive parser,
 \item type class instances, and
 \item combinators.
\end{itemize}

@@

We also want parsers to return a value, if they succeed.  To do this, we extend the return type
to allow for a value, which also requires a second type parameter:
\begin{verbatim}
newtype Parser t a = 
    Parser {getParser :: [t] -> ([t], a)}
\end{verbatim}

However, this ignores the possibility of failure.  To allow for failure, the return value needs to
be wrapped in a Maybe:
\begin{verbatim}
newtype Parser t a = 
    Parser {getParser :: [t] -> Maybe ([t], a)}
\end{verbatim}

@@

The \verb+<*>+ operator of Applicative runs parsers sequentially
and applies the result of the first to the result of the second; it
has to make sure that the (possibly modified) output token stream from
the first parser is passed to the second:
\begin{verbatim}
import Control.Applicative  (Applicative(..))

@@

We can run parsers in sequence using the Applicative combinators.  First, one parser
is run, then the second is run, and the token position is threaded between the two.
Note that both parsers must succeed in order for the combined parser to succeed:
\begin{verbatim}
*Main> getParser (item *> item) "abcde"
Just ("cde",'b')

@@

parseNest :: Parser Char [Nesting]
parseNest = many element
\end{verbatim}
Here are some examples of this code in action:
\begin{verbatim}
*Main> getParser parseNest "(((()))))"
Just (")",[Many [Many [Many [Many []]]]])

@@

*Main> getParser parseNest "(()abc(def)()"
Just ("(()abc(def)()",[])
\end{verbatim}
This concludes the implementation of a simple parser.
In the rest of this article, we'll rework and augment these basic parser 
combinators by generalizing the implementations and extending the result types,
while taking advantage of monad transformers to keep them modular.

@@

This is identical to the Parser type given before modulo the order 
of the result tuple, so we can redefine the parser type as:
\begin{verbatim}
import Control.Monad.State       (StateT(..))

@@

There are two major benefits to our parsers from using monad transformers:
\begin{itemize}
 \item compositional implementation with semantic building blocks --
    the mtl library provides implementations
    for Maybe, Error, and State building blocks; using these, we can easily
    describe the semantics of our parsers without writing very much code, and
 \item type class instances -- in addition to providing the building blocks,
    the mtl provides instances for common type classes, so we only have to 
    implement parser-specific operations.
\end{itemize}

@@

The rest of this articule will progressively adding features to a simple 
initial implementation on our way to creating an error-reporting parser.
The simple language we'll use as a motivation for this task is Woof, a 
simple dialect of Lisp.

The language definition of Woof in pseudo-BNF~\cite{bnf} is:

@@

There is an additional rule that whitespace and comments may appear in any 
amount before or after tokens.  Our tokens are:
\begin{itemize}
  \item \verb+{+
  \item \verb+}+
  \item \verb+(+
  \item \verb+)+
  \item \verb+Symbol+
\end{itemize}

@@

\subsection{Preliminaries}
Our first Woof parser will only be responsible for determining whether an input
text conforms to the language definition and building an 
Abstract Syntax Tree (AST) representing the structure of the parsed input.
Here's the AST definition we'll use:
\begin{verbatim}
data AST
    = ASymbol String
    | ALambda [String] [AST]
    | ADefine String AST
    | AApp    AST  [AST]
  deriving (Show, Eq)
\end{verbatim}

We'll also want a function for running our parsers, using the type given in the
prevous section:
\begin{verbatim}
runParser :: Parser t a -> [t] -> Maybe (a, [t])
runParser = runStateT
\end{verbatim}

@@

Because we used MonadState's combinators to build \verb+item+, \verb+literal+ 
has a slightly different type; we also take advantage of \verb+satisfy+ to 
simplify its implementation:
\begin{verbatim}
{-# LANGUAGE NoMonomorphismRestriction #-}

literal :: (Eq t, MonadState [t] m, Alternative m) => t -> m t
literal c = satisfy ((==) c)
\end{verbatim}

@@

Recall, however, that our specification said that comments and whitespace can occur 
anywhere before or after tokens -- how should this be handled?

First, we need to define the whitespace and comment patterns -- we'll also rename
the \verb+some+ and \verb+many+ functions from Control.Alternative to more
clearly indicate their semantics:
\begin{verbatim}
many0 = many
many1 = some

@@

We want to throw away comments and whitespace, and this process
must occur after every token.  We can do this by using
a combinator which first parses a token and then throws away junk:
\begin{verbatim}
junk = many0 (whitespace <|> comment)

@@

With \verb+tok+ in hand, we can fix \verb+opencurly+:
\begin{verbatim}
opencurly = tok $ literal '{'

@@

Excellent!  The \verb+opencurly+ parser matches an open curly brace, failing
otherwise, and discards trailing whitespace and comments.

@@

The parser for our last token, Symbol, is more complicated, as any number of
alphabetical characters is valid.  We'll use \verb+satisfy+ and a
predicate to check that a character is valid, and \verb+many1+ to 
match one or more valid characters:
\begin{verbatim}
symbol = tok $ many1 char
  where char = satisfy (flip elem (['a' .. 'z'] ++ ['A' .. 'Z']))

@@

The first combining form is function application.  The rule says that it is delimited
by matching parentheses, in between which must appear one form as the operator, followed
by any number of forms as the arguments to which the operator is applied.  We can
say that like so:
\begin{verbatim}
application =
    openparen    *>
    pure AApp   <*>
    form        <*>
    many0 form  <*
    closeparen
\end{verbatim}

@@

\section{Example 2: error-reporting}
While our first parser worked great on valid input, it wasn't helpful
when the input was malformed.  When the parser finds bad input, we'd like it to 
produce an informative error, indicating what and where the problem was.

@@

While our previous parsers had two different results -- Nothing and Just --
our new parsers have three possible results:  success (Right Just), failure (Right Nothing), and error (Left).
%% EZY
A parse failure means that a parser wasn't able to match the input, but
nothing bad happened; failures can be recovered from if there is an alternative
parse that succeeds.  However, an error means that an unrecoverable problem
has been found and that parsing should immediately stop; the parser will
not be able to recover even if there are alternatives, since backtracking
will not be performed.
Note that successful and failing parses are wrapped in an additional \verb+Right+
constructor:
\begin{verbatim}
*Main> runParser item "abc"
Right (Just ('a',"bc"))
